---
title: "整頓資料"
author: "陳紹慶"
date: "2021/4/28"
output: 
  html_document:
    css: custom.css
    toc: true
---

要實施分析計畫之前，必須先整頓為要分析的單位。整頓包括**篩選**有需要的變數或觀察值，以及進行繪圖或重要分析之前的**加工**。

# 學習目標

- `dplyr`總和變數的函式系列
- `dplyr`變數加工的函式系列
- `dplyr`新造變數的函式系列
- 運用字串、日期等字符資料整頓資料表單

- 推薦秘訣表：[Data Import](https://raw.githubusercontent.com/rstudio/cheatsheets/master/data-import.pdf), [Data Transformation](https://raw.githubusercontent.com/rstudio/cheatsheets/master/data-transformation.pdf), [Regular Expressions](https://raw.githubusercontent.com/rstudio/cheatsheets/master/strings.pdf), [Dates and Times](https://raw.githubusercontent.com/rstudio/cheatsheets/master/lubridate.pdf)


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lubridate)
library(dataskills)
set.seed(8675309) # makes sure random numbers are reproducible
```


## 預覽示範資料 `disgust` {#data-disgust}

These examples will use data from [disgust.csv](/data/disgust.csv), which contains data from the [Three Domain Disgust Scale](http://digitalrepository.unm.edu/cgi/viewcontent.cgi?article=1139&context=psy_etds). Each participant is identified by a unique `user_id` and each questionnaire completion has a unique `id`.

```{r, message = FALSE}
disgust <- read_csv("https://psyteachr.github.io/msc-data-skills/data/disgust.csv")

glimpse(disgust)
```

*Questionnaire Instructions*: The following items describe a variety of concepts. Please rate how disgusting you find the concepts described in the items, where 0 means that you do not find the concept disgusting at all, and 6 means that you find the concept extremely disgusting.

| colname   | question                                                                          |
|----------:|:----------------------------------------------------------------------------------|
| moral1 	  | Shoplifting a candy bar from a convenience store                                  |
| moral2	  | Stealing from a neighbor                                                          |
| moral3	  | A student cheating to get good grades                                             |
| moral4	  | Deceiving a friend                                                                |
| moral5	  | Forging someone's signature on a legal document                                   |
| moral6	  | Cutting to the front of a line to purchase the last few tickets to a show         |
| moral7	  | Intentionally lying during a business transaction                                 |
| sexual1	  | Hearing two strangers having sex                                                  |
| sexual2	  | Performing oral sex                                                               |
| sexual3	  | Watching a pornographic video                                                     |
| sexual4	  | Finding out that someone you don't like has sexual fantasies about you            |
| sexual5	  | Bringing someone you just met back to your room to have sex                       |
| sexual6	  | A stranger of the opposite sex intentionally rubbing your thigh in an elevator    |
| sexual7	  | Having anal sex with someone of the opposite sex                                  |
| pathogen1	| Stepping on dog poop                                                              |
| pathogen2	| Sitting next to someone who has red sores on their arm                            |
| pathogen3	| Shaking hands with a stranger who has sweaty palms                                |
| pathogen4	| Seeing some mold on old leftovers in your refrigerator                            |
| pathogen5	| Standing close to a person who has body odor                                      |
| pathogen6	| Seeing a cockroach run across the floor                                           |
| pathogen7	| Accidentally touching a person's bloody cut                                       |
## `dplyr`總和變數的函式系列

Most of the data wrangling you'll want to do with psychological data will involve the `tidyr` verbs you learned in [Chapter 3](#tidyr) and the six main `dplyr` verbs: `select`, `filter`, `arrange`, `mutate`, `summarise`, and `group_by`.

### select() {#select}

選出需要變數、建立小型資料集合

Select columns by name or number.

You can select each column individually, separated by commas (e.g., `col1, col2`). You can also select all columns between two columns by separating them with a colon (e.g., `start_col:end_col`).


```{r}
moral <- disgust %>% select(user_id, moral1:moral7)
names(moral)
```

You can select columns by number, which is useful when the column names are long or complicated.

```{r}
sexual <- disgust %>% select(2, 11:17)
names(sexual)
```

You can use a minus symbol to unselect columns, leaving all of the other columns. If you want to exclude a span of columns, put parentheses around the span first (e.g., `-(moral1:moral7)`, not `-moral1:moral7`).

```{r}
pathogen <- disgust %>% select(-id, -date, -(moral1:sexual7))
names(pathogen)
```

You can select columns based on criteria about the column names.{#select_helpers}

`select()`的好用參數

#### `starts_with()` {#starts_with}

Select columns that start with a character string.

```{r}
u <- disgust %>% select(starts_with("u"))
names(u)
```

#### `ends_with()` {#ends_with}

Select columns that end with a character string.

```{r}
firstq <- disgust %>% select(ends_with("7"))
names(firstq)
```

#### `contains()` {#contains}

Select columns that contain a character string.

```{r}
pathogen <- disgust %>% select(contains("pathogen"))
names(pathogen)
```

#### `num_range()` {#num_range}

Select columns with a name that matches the pattern `prefix`.

```{r}
moral2_4 <- disgust %>% select(num_range("moral", 2:4))
names(moral2_4)
```

```{block, type="info"}
Use `width` to set the number of digits with leading
zeros. For example, `num_range('var_', 8:10, width=2)` selects columns `var_08`, `var_09`, and `var_10`.
```

### filter() {#filter}

選出符合需要的觀察值、建立小型資料集合

Select rows by matching column criteria.

Select all rows where the user_id is 1 (that's Lisa). 

```{r}
disgust %>% filter(user_id == 1)
```

```{block, type="warning"}
Remember to use `==` and not `=` to check if two things are equivalent. A single `=` assigns the righthand value to the lefthand variable and (usually) evaluates to `TRUE`.
```

You can select on multiple criteria by separating them with commas.

多重條件篩選

```{r}
amoral <- disgust %>% filter(
  moral1 == 0, 
  moral2 == 0,
  moral3 == 0, 
  moral4 == 0,
  moral5 == 0,
  moral6 == 0,
  moral7 == 0
)
```

You can use the symbols `&`, `|`, and `!` to mean "and", "or", and "not". You can also use other operators to make equations.

```{r}
# everyone who chose either 0 or 7 for question moral1
moral_extremes <- disgust %>% 
  filter(moral1 == 0 | moral1 == 7)

# everyone who chose the same answer for all moral questions
moral_consistent <- disgust %>% 
  filter(
    moral2 == moral1 & 
      moral3 == moral1 & 
      moral4 == moral1 &
      moral5 == moral1 &
      moral6 == moral1 &
      moral7 == moral1
  )


# everyone who did not answer 7 for all 7 moral questions
moral_no_ceiling <- disgust %>%
  filter(moral1+moral2+moral3+moral4+moral5+moral6+moral7 != 7*7)

```

#### 匹配運算元

Sometimes you need to exclude some participant IDs for reasons that can't be described in code. the `%in%` operator is useful here for testing if a column value is in a list. Surround the equation with parentheses and put `!` in front to test that a value is not in the list.

**運用匹配運算元 %in% 設定篩選條件**

```{r}
length(unique(disgust$user_id))

no_researchers <- disgust %>%
  filter(!(user_id %in% c(1,2)))

glimpse(no_researchers)
```

#### Dates

資料數值有日期時間時，使用`lubridate`的函式能事半功倍

You can use the `lubridate` package to work with dates. For example, you can use the `year()` function to return just the year from the `date` column and then select only data collected in 2010.

```{r}
disgust2010 <- disgust  %>%
  filter(year(date) == 2010)

glimpse(disgust2010)
```

Or select data from at least 5 years ago. You can use the `range` function to check the minimum and maxiumum dates in the resulting dataset.

```{r}
disgust_5ago <- disgust %>%
  filter(date < today() - dyears(5))

range(disgust_5ago$date)
```


### arrange() {#arrange}

資料按照指定變數數值重新排序

Sort your dataset using `arrange()`.

```{r}
disgust_order <- disgust %>%
  arrange(id)

head(disgust_order)
```

Reverse the order using `desc()`

```{r}
disgust_order <- disgust %>%
  arrange(desc(id))

head(disgust_order)
```


### mutate() {#mutate}

以原變數數值為內容重新組合計算，造出分析需要的新變數。

Add new columns. This is one of the most useful functions in the tidyverse.

Refer to other columns by their names (unquoted). You can add more than one column, just separate the columns with a comma. Once you make a new column, you can use it in further column definitions e.g., `total` below).

```{r}
disgust_total <- disgust %>%
  mutate(
    pathogen = pathogen1 + pathogen2 + pathogen3 + pathogen4 + pathogen5 + pathogen6 + pathogen7,
    moral = moral1 + moral2 + moral3 + moral4 + moral5 + moral6 + moral7,
    sexual = sexual1 + sexual2 + sexual3 + sexual4 + sexual5 + sexual6 + sexual7,
    total = pathogen + moral + sexual,
    user_id = paste0("U", user_id)
  )
```

```{block, type="warning"}
You can overwrite a column by giving a new column 
the same name as the old column. Make sure that you mean to do this and that you aren't trying to use the old column value after you redefine it.
```

### summarise() {#summarise}

指定變數進行需要的描述統計，輸出到新的資料集合。

Create summary statistics for the dataset. Check the [Data Wrangling Cheat Sheet](https://www.rstudio.org/links/data_wrangling_cheat_sheet) or the [Data Transformation Cheat Sheet](https://github.com/rstudio/cheatsheets/raw/master/source/pdfs/data-transformation-cheatsheet.pdf) for various summary functions. Some common ones are: `mean()`, `sd()`, `n()`, `sum()`, and `quantile()`.

```{r}
disgust_total %>%
  summarise(
    n = n(),
    q25 = quantile(total, .25, na.rm = TRUE),
    q50 = quantile(total, .50, na.rm = TRUE),
    q75 = quantile(total, .75, na.rm = TRUE),
    avg_total = mean(total, na.rm = TRUE),
    sd_total  = sd(total, na.rm = TRUE),
    min_total = min(total, na.rm = TRUE),
    max_total = max(total, na.rm = TRUE)
  )
```


### group_by() {#group_by}

指定變數為分組檢索值。

Create subsets of the data. You can use this to create summaries, 
like the mean value for all of your experimental groups.

Here, we'll use `mutate` to create a new column called `year`, group by `year`, and calculate the average scores.

```{r}
disgust_total %>%
  mutate(year = year(date)) %>%
  group_by(year) %>%
  summarise(
    n = n(),
    avg_total = mean(total, na.rm = TRUE),
    sd_total  = sd(total, na.rm = TRUE),
    min_total = min(total, na.rm = TRUE),
    max_total = max(total, na.rm = TRUE)
  )
```

You can use `filter` after `group_by`. The following example returns the lowest total score from each year.

```{r}

disgust_total %>%
  mutate(year = year(date)) %>%
  select(user_id, year, total) %>%
  group_by(year) %>%
  filter(rank(total) == 1) %>%
  arrange(year)

```

You can also use `mutate` after `group_by`. The following example calculates subject-mean-centered scores by grouping the scores by `user_id` and then subtracting the group-specific mean from each score. <span class="text-warning">Note the use of `gather` to tidy the data into a long format first.</span>

```{r}

disgust_smc <- disgust %>%
  gather("question", "score", moral1:pathogen7) %>%
  group_by(user_id) %>%
  mutate(score_smc = score - mean(score, na.rm = TRUE))

```


### All Together

在此示範清理到整頓資料一次完成。

A lot of what we did above would be easier if the data were tidy, so let's do that first. Then we can use `group_by` to calculate the domain scores.

```{block, type="warning"}
It is good practice to use `ungroup()` after using `group_by` and `summarise`. Forgetting to ungroup the dataset won't affect some further processing, but can really mess up other things.
```

Then we can spread out the 3 domains, calculate the total score, remove any rows with a missing (`NA`) total, and calculate mean values by year.

```{r}
## 計算每位參與者在各向度的平均評分
disgust_tidy <- read_csv("../data/disgust.csv") %>%
  gather("question", "score", moral1:pathogen7) %>%
  separate(question, c("domain","q_num"), sep = -1) %>%
  group_by(id, user_id, date, domain) %>%
  summarise(score = mean(score)) %>%
  ungroup() 

head(disgust_tidy)


disgust_tidy2 <- disgust_tidy %>%
  spread(domain, score) %>%
  mutate(
    total = moral + sexual + pathogen, ## 計算三個向度的平均總分
    year = year(date)   ## 製造新變數year儲存年份資訊
  ) %>%
  filter(!is.na(total)) %>%  ## 排除沒有答完所有題目的參與者
  arrange(user_id) 

head(disgust_tidy2)

disgust_tidy3 <- disgust_tidy2 %>%
  group_by(year) %>%       ## 依年分計算人次、各向度平均評分、第一位及最後一位參與者id
  summarise(
    n = n(),
    avg_pathogen = mean(pathogen),
    avg_moral = mean(moral),
    avg_sexual = mean(sexual),
    first_user = first(user_id),
    last_user = last(user_id)
  )

head(disgust_tidy3)

```

## `dplyr`變數加工的函式系列

Use the code examples below and the help pages to figure out what the following one-table verbs do. Most have pretty self-explanatory names.

### rename() {#rename}

為變數改名字

```{r rename}
iris_underscore <- iris %>%
  rename(sepal_length = Sepal.Length,
         sepal_width = Sepal.Width,
         petal_length = Petal.Length,
         petal_width = Petal.Width)

names(iris_underscore)
```

```{block type="try"}
Almost everyone gets confused at some point with `rename()` and tries to put the original names on the left and the new names on the right. Try it and see what the error message looks like.
```

### distinct() {#distinct}

移除重覆的觀察值

```{r distinct}
# create a data table with duplicated values
dupes <- tibble(
  id = rep(1:5, 2),
  dv = rep(LETTERS[1:5], 2)
)

distinct(dupes)
```

### count() {#count}

依指定變數，計算觀察值個數

```{r count}
# how many observations from each species are in iris?
count(iris, Species)
```


### slice() {#slice}

移除指定列的觀察值

```{r slice}

tibble(
  id = 1:10,
  condition = rep(c("A","B"), 5)
) %>%
  slice(3:6, 9)

```


### pull() {#pull}

抽出指定的變數

```{r pull}

iris %>%
  group_by(Species) %>%
  summarise_all(mean) %>%
  pull(Sepal.Length)

```


## `dplyr`新造變數的函式系列 {#window}

必須置於總和變數函式的參數，才能發揮作用。

Window functions use the order of rows to calculate values. You can use them to do things that require ranking or ordering, like choose the top scores in each class, or acessing the previous and next rows, like calculating cumulative sums or means.

The [dplyr window functions vignette](https://dplyr.tidyverse.org/articles/window-functions.html) has very good detailed explanations of these functions, but we've described a few of the most useful ones below. 

### Ranking functions

`rank`系列函數計算指定變數內數值的排序，輸出到指定新變數。

```{r}
tibble(
  id = 1:5,
  "Data Skills" = c(16, 17, 17, 19, 20), 
  "Statistics"  = c(14, 16, 18, 18, 19)
) %>%
  gather(class, grade, 2:3) %>%
  group_by(class) %>%
  mutate(row_number = row_number(),
         rank       = rank(grade),
         min_rank   = min_rank(grade),
         dense_rank = dense_rank(grade),
         quartile   = ntile(grade, 4),
         percentile = ntile(grade, 100))
```

```{block type="try"}
* What are the differences among `row_number()`, `rank()`, `min_rank()`, `dense_rank()`, and `ntile()`? 
* Why doesn't `row_number()` need an argument? 
* What would happen if you gave it the argument `grade` or `class`? 
* What do you think would happen if you removed the `group_by(class)` line above? 
* What if you added `id` to the grouping?
* What happens if you change the order of the rows?
* What does the second argument in `ntile()` do?
```

You can use window functions to group your data into quantiles.

使用`rank`系列函數分組，再做其他整頓工作。

```{r}
iris %>%
  group_by(tertile = ntile(Sepal.Length, 3)) %>%
  summarise(mean.Sepal.Length = mean(Sepal.Length))

```

### Offset functions

依指定變數內數值的登錄順序，紀錄前x位(`lag()`)或後x位(`lead()`)的數值，輸出到新變數

```{r}
lag_lead <- tibble(x = 1:6) %>%
  mutate(lag = lag(x, default = 0),
         lag2 = lag(x, n = 2),
         lead = lead(x, default = 0))
head(lag_lead)
```


```{r}
trials <- tibble(
  trial = sample(1:10, 10),
  cond = sample(c("exp", "ctrl"), 10, T),
  score = rpois(10, 4)
) %>%
  mutate(
    score_change = score - lag(score, order_by = trial),
    change_cond = cond != lag(cond, order_by = trial, 
                              default = "no condition")
  ) %>%
  arrange(trial)
```

```{block type="try"}
Look at the help pages for `lag()` and `lead()`.

* What happens if you remove the `order_by` argument or change it to `cond`?
* What does the `default` argument do?
* Can you think of circumstances in your own data where you might need to use `lag()` or `lead()`?
```

### Cumulative aggregates

依指定變數內數值的登錄順序，逐項計算並輸出到新變數。

`cumsum()`, `cummin()`, and `cummax()` are base R functions for calcumaling cumulative means, minimums, and maximums. The dplyr package introduces `cumany()` and `cumall()`, which return `TRUE` if any or all of the previous values meet their criteria.

```{r}
tibble(
  time = 1:10,
  obs = c(1, 0, 1, 2, 4, 3, 1, 0, 3, 5)
) %>%
  mutate(
    cumsum = cumsum(obs),
    cummin = cummin(obs),
    cummax = cummax(obs),
    cumany = cumany(obs == 3),
    cumall = cumall(obs < 4)
  )
```

```{block type="try"}
* What would happen if you change `cumany(obs == 3)` to `cumany(obs > 2)`?
* What would happen if you change `cumall(obs < 4)` to `cumall(obs < 2)`?
* Can you think of circumstances in your own data where you might need to use `cumany()` or `cumall()`?
```
